---
title: "read_glodap"
author: "Jens Daniel M체ller"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---
---
title: "GLODAPv2_2021"
author: "Jens Daniel M체ller"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  workflowr::wflow_html:
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
editor_options:
  chunk_output_type: console
---

```{r parent, child = "/nfs/kryo/work/jenmueller/emlr_cant/utilities/setup.Rmd"}
# this chunk runs the code stored in setup.Rmd
# if required, please refer to instructions given here:
# https://jdblischak.github.io/workflowr/articles/wflow-07-common-code.html
```

```{r define_paths}

path_glodapv2_2023  <- "/nfs/kryo/work/datasets/ungridded/3d/ocean/glodap/v2023"

```

```{r load_libraries_specific, include=FALSE}
library(lubridate)
library(ggrepel)
library(kableExtra)
```

# Read files

## Adjusted data

Main data source for this project is `GLODAPv2.2023_Merged_Master_File.csv` downloaded from `https://www.ncei.noaa.gov/data/oceans/ncei/ocads/data/0283442/GLODAPv2.2023_Merged_Master_File.csv` on Sep 26, 2024.

```{r read_GLODAPv2_2023_merged_master_file}

GLODAP <-
  read_csv(
    paste(
      path_glodapv2_2023,
      "GLODAPv2.2023_Merged_Master_File.csv",
      sep = "/"
    ),
    na = "-9999",
    col_types = cols(.default = col_double())
  )


GLODAP <- GLODAP %>%
  rename_with(~str_remove(., 'G2'))

GLODAP <- GLODAP %>% 
  select(expocode:bottomdepth,
         depth, temperature,
         contains(c("salinity", "talk", "tco2")))


GLODAP <- GLODAP %>% 
  filter(!is.na(talk),
         !is.na(salinity),
         bottomdepth > 1000,
         talkf == 2,
         talkqc == 1,
         salinityf == 2,
         salinityqc == 1)


```



## Harmonize nomenclature

```{r harmonize_variables}

# create date column
GLODAP <- GLODAP %>%
  mutate(date = ymd(paste(year, month, day))) %>%
  relocate(date)

# harmonize column names
GLODAP <- GLODAP  %>%
  rename(sal = salinity,
         temp = temperature)

# harmonize coordinates
GLODAP <- GLODAP  %>%
  rename(lon = longitude,
         lat = latitude) %>%
  mutate(lon = if_else(lon < 20, lon + 360, lon))



```


## Horizontal gridding

For merging with other data sets, all observations were grouped into latitude intervals of:

-   1째 x 1째

```{r grid_spatially_1x1}

GLODAP <- m_grid_horizontal(GLODAP)
GLODAP <- m_grid_horizontal_coarse(GLODAP)

```


```{r talk_trends, fig.asp=1.5}


GLODAP <- GLODAP  %>%
  mutate(stalk = talk * 35 / sal)

TA_trend <- GLODAP %>%
  mutate(lat = lat_grid, lon = lon_grid) %>% 
  mutate(depth_layer = cut(depth, seq(0, 2000, 50))) %>%
  filter(!is.na(depth_layer)) %>%
  filter(year >= 1990) %>%
  group_by(year, lon, lat, depth_layer) %>%
  summarise(stalk = mean(stalk, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = depth_layer, values_from = stalk) %>%
  pivot_longer(-c(1:4), names_to = "reference_depth") %>%
  mutate(surface_deep = `(0,50]` - value) %>%
  drop_na() %>%
  group_by(lon, lat, reference_depth) %>%
  mutate(n = n(), delta_year = max(year) - min(year)) %>%
  ungroup() %>%
  filter(n >= 2, delta_year >= 10) %>%
  group_by(reference_depth) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n >= 0.5 * max(n)) %>%
  select(year, lon, lat, surface_deep, reference_depth) %>%
  nest(data = -c(lon, lat, reference_depth)) %>%
  mutate(fit = map(data, ~ flm(
    formula = surface_deep ~ year, data = .x
  ))) %>%
  unnest_wider(fit) %>%
  select(lon, lat, reference_depth, slope = year) %>%
  mutate(slope = as.vector(slope))


TA_trend %>% 
  filter(abs(slope) < 1) %>%
  ggplot(aes(slope)) +
  geom_vline(xintercept = 0) +
  geom_histogram() +
  facet_wrap(~ reference_depth, ncol = 3)

TA_trend %>% 
  filter(abs(slope) < 1) %>%
  ggplot(aes(lon, lat, fill = slope)) +
  geom_tile() +
  coord_quickmap() +
  scale_fill_divergent(mid = "grey") +
  facet_wrap(~ reference_depth, ncol = 3)
  

```


```{r tco2_trends, fig.asp=1.5}


GLODAP <- GLODAP  %>%
  mutate(stco2 = tco2 * 35 / sal)

TA_trend <- GLODAP %>%
  mutate(lat = lat_grid, lon = lon_grid) %>% 
  mutate(depth_layer = cut(depth, seq(0, 2000, 50))) %>%
  filter(!is.na(depth_layer)) %>%
  filter(year >= 1990) %>%
  group_by(year, lon, lat, depth_layer) %>%
  summarise(stco2 = mean(stco2, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = depth_layer, values_from = stco2) %>%
  pivot_longer(-c(1:4), names_to = "reference_depth") %>%
  mutate(surface_deep = `(0,50]` - value) %>%
  drop_na() %>%
  group_by(lon, lat, reference_depth) %>%
  mutate(n = n(), delta_year = max(year) - min(year)) %>%
  ungroup() %>%
  filter(n >= 2, delta_year >= 10) %>%
  group_by(reference_depth) %>%
  mutate(n = n()) %>%
  ungroup() %>%
  filter(n >= 0.5 * max(n)) %>%
  select(year, lon, lat, surface_deep, reference_depth) %>%
  nest(data = -c(lon, lat, reference_depth)) %>%
  mutate(fit = map(data, ~ flm(
    formula = surface_deep ~ year, data = .x
  ))) %>%
  unnest_wider(fit) %>%
  select(lon, lat, reference_depth, slope = year) %>%
  mutate(slope = as.vector(slope))


TA_trend %>% 
  filter(abs(slope) < 5) %>%
  ggplot(aes(slope)) +
  geom_vline(xintercept = 0) +
  geom_histogram() +
  facet_wrap(~ reference_depth, ncol = 3)

TA_trend %>% 
  filter(abs(slope) < 5) %>%
  ggplot(aes(lon, lat, fill = slope)) +
  geom_tile() +
  coord_quickmap() +
  scale_fill_divergent(mid = "grey") +
  facet_wrap(~ reference_depth, ncol = 3)
  

```

